---
title: "Using GIS environmental data for gender statistics analysis"
subtitle: "A step-by-step case study with Bangladesh data "
author: "Christophe Bontemps & Eunkoo Lee (SIAP)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: lumen
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( message = FALSE, warning = FALSE, results =FALSE, echo = TRUE) 
```

* * *
# 1. Introduction to integrating household survey and geospatial data

## GIS data sources

There are many data sources freely available with environmental information at a very detailed level. These files are from huge data bases that cover large areas of the word. 

> Some sources may require a login and an enrollment to some institution. Most of the files are very "heavy". 

The table hereafter provides some links to the resources we have used for this analysis. 


|     Geo-covariates      |     Definition                                                                                                                                               |     Data   source link (use Google Chrome)                                                                                                |
|-------------------------|--------------------------------------------------------------------------------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------|
|     Travel_Times2015    |     Travelling time (in minutes) to the nearest city of   more than 50,000 people                                                                            |     https://doi.org/10.6084/m9.figshare.7638134.v3                                                                                        |
|     SMOD2015            |     Degree of urbanization                                                                                                                                   |     http://cidportal.jrc.ec.europa.eu/ftp/jrc-opendata/GHSL/GHS_SMOD_POP_GLOBE_R2016A/                                                    |
|     Buildup2015         |     Percentage of building footprint area in relation to   the total cell area.                                                                              |     http://cidportal.jrc.ec.europa.eu/ftp/jrc-opendata/GHSL/GHS_BUILT_LDSMT_GLOBE_R2015B/GHS_BUILT_LDS2014_GLOBE_R2016A_54009_1k/V1-0/    |
|     Aridity2015         |     Climate data related to evapotranspiration processes   and rainfall deficit for potential vegetative growth. Higher index suggests   higher humidity.    |     https://figshare.com/articles/Global_Aridity_Index_and_Potential_Evapotranspiration_ET0_Climate_Database_v2/7504448/3                 |
|     Density2015         |     Number of inhabitants per   cell (1km X 1km)                                                                                                             |     http://cidportal.jrc.ec.europa.eu/ftp/jrc-opendata/GHSL/GHS_POP_GPW4_GLOBE_R2015A/GHS_POP_GPW42015_GLOBE_R2015A_54009_1k/V1-0/        |
|     aIncome2013         |     Estimates of income in USD   per grid square                                                                                                             |     https://www.worldpop.org/doi/10.5258/SOTON/WP00020                                                                                    |
|     aPP2013             |     Mean likelihood of living in   poverty per grid square                                                                                                   |     https://www.worldpop.org/doi/10.5258/SOTON/WP00020                                                                                    |
|     aWealthIndex2011    |     Mean DHS wealth index score   per grid square                                                                                                            |     https://www.worldpop.org/doi/10.5258/SOTON/WP00020                                                                                    |

### Some good practices for file management {-}

We need a minimum of organisation in the data and code folders, as well as some R packages.

```{r packages}
# GIS packages
library(raster) ## for reading "RASTER" files
library(rgdal)  ## for reading "shapefiles"
library(sp)     

# Tidy data management packages
library(dplyr)
library(data.table)

# Plotting packages
library(ggplot2)
library(RColorBrewer)

# Nice presentation of results
library(knitr)
library(papeR)
```

The *raw* files we have downloaded (and which are heavy) can be stored in a project folder for future usage. We distinguish here several folders depending on the source and type of the data sets. It is also important to use a different folder for the data set we'll be creating. 

```{r}
# --- Change to YOUR project folder HERE  ---- #

# source_folder<-"c:/GitMain/UNWomen-GISBangladesh/"
source_folder<-"c:/Chris/UN-ESCAP/MyCourses/GenderStats/Code/Annex2/"

data_folder<-paste(source_folder,"Data/",sep="")

# Specific sub-folder for data storage
shapedata_folder<-paste(source_folder, "Data/dhsdata/BDGE71FL", sep="")  
geodata_folder<-paste(source_folder, "Data/geodata/", sep="")

# this is where all saved .Rda go and can be loaded when needed
CreatedData<-paste(source_folder, "CreatedData/" , sep="")  
```


# 2. Understanding child marriage using geo-covariates

## The DHS survey 

```{r}
# Reading DHS survey data 
DHS<-read.csv(file = paste0(source_folder,'Data/bangladesh.csv'))  # reading DHS Bangladesh 2014
DHS$Age<-as.numeric(DHS$Age)

#Computing the proportion of getting married before 15 by cluster
cluster_average<-aggregate(Before15~DHSCLUST,
                           data=DHS,
                           FUN=mean)  
```

### Reading the DHS Shapefile {-} 

```{r}
# Reading DHS Shapefile 
dhsShapeData<-readOGR(shapedata_folder, "BDGE71FL") # Reads the shapefile in DHS
shapedata<-dhsShapeData@data                        # Reads the data part 
shapefile_df <- fortify(shapedata)
shapedata<-shapedata[shapedata$LATNUM>0, ]          # Drops negative Latnum 
```

### Cluster locations by urban and rural from the shapefile (DHS Bangladesh 2014) {-}

We can use the *latitude* and *longitude* of each observation  to draw a "location map" of the clusters.  

>This is not really "a map", but only points with latitude and longitude defined represented on a grid


```{r}
# Now the shapefile can be plotted, as points 
# In the aesthetics part of the ggplot we need long, lat,
#                       and we use group for Urban or Rural.
map <- ggplot() +
  geom_point(data = shapedata,
              aes(x = LONGNUM, y = LATNUM, color = factor(URBAN_RURA)),
              size = .6) +
 ggtitle("Shapefile Cluster Location on and X-Y grid") + 
         labs(x=  "X-axis used as Longitude" ,
              y = "Y-axis used as Latitude")

# Using the ggplot2 function coord_map will make things look better
# and it will also let you change the projection.
map_projected <- map +
  coord_map()+
  theme_minimal()

map_projected

```



## Using GIS sources 

### Matching DHS (*Shapefile*) and CIS (*raster*) files using CRS {-} 

We need to know what is the **CRS** (Coordinates Reference System) for the Raster file in order to use the same for the DHS data and to have exact correspondence of the two geographical systems


```{r, results= TRUE}
# Reading geographic data - access to cities
accessData<-raster(paste(geodata_folder,
                         "accessibility_to_cities_2015.tif",
                         sep=""))
accessData
```
Here the CRS is "*CRS = +proj=longlat +datum=WGS84 +no_defs*" so we know it is using the *WGS84*^[WGS 84 is an Earth-centered, Earth-fixed terrestrial reference system and geodetic datum. WGS 84 is based on a consistent set of constants and model parameters that describe the Earth's size and shape (see https://en.wikipedia.org/wiki/World_Geodetic_System.]

> We need to provide the same  projection reference for the DHS data (shapefile)

We can use the package *sp* and the *spTranform* function for that. 

```{r}
#  We use sp::spTransform to tell wich CRS is used.

dhsShapeData <- spTransform(dhsShapeData, accessData@crs)
dhsShapeData

```

### Extracting values from a raster {-}    

We can now **extract** the values from the *acessData* file (a Raster object) at the locations of our household (shapefile). The result is a data frame. The first column is a sequential ID, the other columns are the extracted values, i.e. the travel time to a city for each cluster. 

> Extracting values takes time. 

#### Example with the poverty index (Mean likelihood of living in poverty per grid square - *worldpop.org*) {-}

- **First**, We can just **read** the raster file

```{r RasterReadingAPP, cache = TRUE}
# Reading raster file for aPP2013 
aPPData<-raster(paste(geodata_folder, "bgd2013ppipov.tif", sep=""))
```

-  Then, we can **visualize** the information from this file on a map, here a poverty map. 

```{r}
# Main plot using the plotting function of raster package
plot(aPPData, 
     breaks=c(0, 60, 70, 75, 80, 85, 100, 150),  
     col = terrain.colors(8),
     main="Poverty Map") 
```

- **Second**,  we can **extract** the information from the raster file for each cluster location. This operation, using the *extract* function from the *raster* package^[see the documentation https://www.rdocumentation.org/packages/raster/versions/3.4-13/topics/extract],  takes a lot of resources and **can be lengthy! **

```{r RasterExtractionAPP, cache = TRUE}
dhs_all2000 <- extract(aPPData,         # Raster file
                       dhsShapeData,    # Shapefile with DHS information   
                       buffer = 2000,   # buffer size (meters) around each cluster 
                       df=TRUE)         # returns a data frame

dhs_all2000<-as.data.frame(dhs_all2000) # Because it may not return a data frame
```

We may remove some NAs generated by the matching 

```{r CleaningAPP}
temp<-dhs_all2000[!is.na(dhs_all2000$bgd2013ppipov), ]  # removing NAs in poverty estimates
```

There may be several poverty estimates values in each cluster, so we'll aggregate this information at the cluster level. 

```{r AggregationAPP}
# Aggregating at the cluster level

aPPData.agg<-aggregate(temp$bgd2013ppipov,
                       by=list(temp$ID),
                       FUN=mean)                      
# Adding column name
colnames(aPPData.agg)<-c("DHSCLUST", "aPP2013")
# save(aPPData.agg, file="CreatedData/aPPData.Rda")
```

#### Map of PSU locations on poverty map from the raster file {-}

```{r}
# Main plot using the plotting function of raster package
plot(aPPData, 
     breaks=c(0, 60, 70, 75, 80, 85, 100, 150),  
     col = terrain.colors(8),
     main="Map of PSU locations on Poverty Map", 
     sub="Probability of Poverty") 

# We can add points for each cluster location on this map
points(x=shapedata$LONGNUM, 
       y=shapedata$LATNUM, 
       type="p", 
       cex=0.3, 
       pch=21, 
       bg=1)
```


### Importing other geographical information from other raster files {-}

The exact same operations can be done with all the other geographic files we have identified


> These operations take time and you want to skip these steps and upload directly the file created. 

####  Aridity Index (Global Aridity Index - *wordclim.org*) {-}

```{r RasterExtractionAridity, cache = TRUE, eval =FALSE}
# Reading raster file for Aridity2015 
# memory.limit(9999999999)
# aridityData <- raster(readGDAL(paste(geodata_folder, "AI_annual/ai_yr/w001001.adf", sep="")))
# dhsShapeData2 <- spTransform(dhsShapeData, aridityData@crs)
# dhs_all2000 <- extract(aridityData,
#                        dhsShapeData2,
#                        buffer = 2000,
#                        df=TRUE)
# dhs_all2000<-as.data.frame(dhs_all2000)
# dhs_all2000<-dhs_all2000[!is.na(dhs_all2000$band1),]
# 
#  aridityData.agg<-aggregate(dhs_all2000$band1,
#                             by=list(dhs_all2000$ID),
#                             FUN=mean)
# colnames(aridityData.agg)<-c("DHSCLUST", "Aridity2015")
# #save(aridityData.agg, file="CreatedData/aridityData.Rda")
```

####  Wealth Index (Mean wealth index score - *Worldpop.org* ) {-}

```{r RasterExtractionWealth, cache = TRUE, eval=FALSE}
# Reading raster file for aWealthindex2011 
# aWIData<-raster(paste(geodata_folder, "bgd2011wipov.tif", sep=""))
# dhs_all2000 <- extract(aWIData,    
#                        dhsShapeData,          
#                        buffer = 2000,     
#                        df=TRUE)           
# dhs_all2000<-as.data.frame(dhs_all2000)
# temp<-dhs_all2000[!is.na(dhs_all2000$bgd2011wipov), ]
# 
# aWIData.agg<-aggregate(temp$bgd2011wipov,
#                        by=list(temp$ID),
#                        FUN=mean)
# colnames(aWIData.agg)<-c("DHSCLUST", "aWealthIndex2011")
# #save(aWIData.agg, file="CreatedData/aWIData.Rda")

```

####  Travel Time to Cities Index (Nature - Extracted  from Google Maps Platform Distance Matrix API) {-}
*Travelling time (in minutes) to the nearest city of more than 50,000 people*

```{r RasterExtractionAccess, cache = TRUE}
# Data extraction using the matching between raster and Spatial data
# dhs_all2000 <- raster::extract(accessData,     # raster layer
#                                dhsShapeData2,  # SPDF with centroids for buffer
#                                 buffer = 2000, # buffer size (meters)
#                                 df=TRUE)       # returns a dataframe
#     
# dhs_all2000<-as.data.frame(dhs_all2000)

### Filtering to remove na values and distances equal to 0. 

# dhs_all2000<-dhs_all2000[!is.na(dhs_all2000$accessibility_to_cities_2015)
#                          & dhs_all2000$accessibility_to_cities_2015>=0, ]
# 
# # Aggregation (mean of the travel times for each cluster)
# # Name changed here to avoid erasing row data: acessData --> accessData.agg
# accessData.agg<-aggregate(dhs_all2000$accessibility_to_cities_2015, 
#                           by=list(dhs_all2000$ID), 
#                           FUN=mean)
# colnames(accessData.agg)<-c("DHSCLUST", "Travel_Times2015")
# 
# # Saving the file in a devoted folder 
# save(accessData.agg, file="CreatedData/accessData.Rda")
```

#### Urbanization Index (Degree of urbanization - *Europa.eu*) {-} 

```{r RasterExtractionSmod, cache = TRUE, eval =FALSE}
# !LENGHTY OPERATION!

# # Reading raster file for SMOD2015 
# smodData<-raster(paste(geodata_folder, "GHS_SMOD_POP2015_GLOBE_R2016A_54009_1k_v1_0.tif", sep=""))
# dhsShapeData2 <- spTransform(dhsShapeData, smodData@crs)
# dhs_all2000 <- extract(smodData,    # raster layer
#                        dhsShapeData2,         
#                        buffer = 2000,     
#                        df=TRUE)           
# dhs_all2000<-as.data.frame(dhs_all2000) 
# 
# smodData.agg<-aggregate(dhs_all2000$GHS_SMOD_POP2015_GLOBE_R2016A_54009_1k_v1_0,
#                         by=list(dhs_all2000$ID),
#                         FUN=mean)
# colnames(smodData.agg)<-c("DHSCLUST", "SMOD2015")
# save(smodData.agg, file="CreatedData/smodData.Rda")
```


#### Building concentration index (Percentage of building - * European Union*) {-} 
 
```{r RasterExtractionBuildup, cache = TRUE, eval =FALSE}
# # Reading raster file for Buildup2015 
# buildupData<-raster(paste(geodata_folder, "GHS_BUILT_LDS2014_GLOBE_R2016A_54009_1k_v1_0.tif", sep=""))
# dhsShapeData2 <- spTransform(dhsShapeData, buildupData@crs)
# dhs_all2000 <- extract(buildupData,    
#                        dhsShapeData2,          
#                        buffer = 2000,    
#                        df=TRUE)           
# dhs_all2000<-as.data.frame(dhs_all2000)
# 
# buildupData.agg<-aggregate(dhs_all2000$GHS_BUILT_LDS2014_GLOBE_R2016A_54009_1k_v1_0,
#                            by=list(dhs_all2000$ID),
#                            FUN=mean)
# colnames(buildupData.agg)<-c("DHSCLUST", "Buildup2015")
# save(buildupData.agg, file="CreatedData/buildupData.Rda")
```

#### Density Index (Number of inhabitants per cell (1km X 1km) - *European Union* ) {-} 

```{r RasterExtractionDensity, cache = TRUE, eval =FALSE}
# # Reading raster file for Density2015 
# densityData<-raster(paste(geodata_folder, "GHS_POP_GPW42015_GLOBE_R2015A_54009_1k_v1_0.tif", sep=""))
# dhsShapeData2 <- spTransform(dhsShapeData, densityData@crs)
# dhs_all2000 <- extract(densityData,    
#                        dhsShapeData2,          
#                        buffer = 2000,     
#                        df=TRUE)          
# dhs_all2000<-as.data.frame(dhs_all2000)
# 
# densityData.agg<-aggregate(dhs_all2000$GHS_POP_GPW42015_GLOBE_R2015A_54009_1k_v1_0,
#                            by=list(dhs_all2000$ID),
#                            FUN=mean)
# colnames(densityData.agg)<-c("DHSCLUST", "Density2015")
# save(densityData.agg, file="CreatedData/densityData.Rda")
```

#### Income Index (Estimates of income in USD per grid square - *Worldpop.org*) {-} 

```{r RasterExtractionIncome, cache = TRUE, eval =FALSE}
# # Reading raster file for aIncome2013 
# aICData<-raster(paste(geodata_folder, "bgd2013incpov.tif", sep=""))
# dhs_all2000 <- extract(aICData,    
#                        dhsShapeData,         
#                        buffer = 2000,     
#                        df=TRUE)          
# dhs_all2000<-as.data.frame(dhs_all2000)
# temp<-dhs_all2000[!is.na(dhs_all2000$bgd2013incpov), ]
# 
# aICData.agg<-aggregate(temp$bgd2013incpov,
#                        by=list(temp$ID),
#                        FUN=mean)
# colnames(aICData.agg)<-c("DHSCLUST", "aIncome2013")
# save(aICData.agg, file="CreatedData/aICData.Rda")
```


# 4. Logistic regression and Random Forests 

The DHS survey records individual information on various aspects and 


## Research question

Child marriage (marriage before the age of 18) is a fundamental violation of human rights. Girls who marry before they turn 18 are less likely to remain in school and more likely to experience domestic violence. Despite laws against child marriage, 650 million girls and women alive today were married as children. 

> Several factors are known to influence the probability of getting married before 15 or 18. How about **environment factors**? Do environment-related factors such as drought episodes, level of aridity and urbanization have any impact on child marriage? 


#### Technical remark {-}
> Since the previous operations may take time and CPU resources, you can directly load the data sets created above and **start using the code here**

We have stored on SIAp's website the files created in the integration process described above. You can then use these files to proceed with the analysis.   

```{r FileLoading, cache = TRUE}
### Loading  Geo-covariate for clusters from SIAP's server ## 

load(url("https://www.unsiap.or.jp/on_line/Big_Data/accessData.Rda"))
load(url("https://www.unsiap.or.jp/on_line/Big_Data/smodData.Rda"))
load(url("https://www.unsiap.or.jp/on_line/Big_Data/buildupData.Rda"))
load(url("https://www.unsiap.or.jp/on_line/Big_Data/aridityData.Rda"))
load(url("https://www.unsiap.or.jp/on_line/Big_Data/densityData.Rda"))
load(url("https://www.unsiap.or.jp/on_line/Big_Data/aWIData.Rda"))
load(url("https://www.unsiap.or.jp/on_line/Big_Data/aICData.Rda"))
load(url("https://www.unsiap.or.jp/on_line/Big_Data/aPPData.Rda"))
```

For our analysis, we will first merge all the geospatial  information into a simple data frame. We create a function (*dhsdataMerge*) for that.  

```{r}
## Function used for merging geo-covariates to DHS data #
dhsdataMerge<-function(originalData){
  datause<-merge(originalData, accessData.agg, by=c("DHSCLUST"), all.x=T)
  datause<-merge(datause, smodData.agg, by=c("DHSCLUST"), all.x=T)
  datause<-merge(datause, buildupData.agg, by=c("DHSCLUST"), all.x=T)
  datause<-merge(datause, aridityData, by=c("DHSCLUST"), all.x=T)  ## NO .agg HERE because you gave it to me already aggregated !!! 
  datause<-merge(datause, densityData.agg, by=c("DHSCLUST"), all.x=T)
  datause<-merge(datause, aWIData.agg, by=c("DHSCLUST"), all.x=T)
  datause<-merge(datause, aICData.agg, by=c("DHSCLUST"), all.x=T)
  datause<-merge(datause, aPPData.agg, by=c("DHSCLUST"), all.x=T)
  datause<-datause[datause$DHSCLUST!=544,]
  return(datause)
}
# Using this function, we can merge the file cluster_average
# with all the Geo-covariate extracted at the cluster level
data.agg<-dhsdataMerge(cluster_average) 
```

### Correlation plot {-}

> It is always a good practice to analyse the relationships between the variables

```{r }
library(ggcorrplot)

# We compute the correlation matrix of the covariates
corr_coef<-cor(data.agg[, c(3:10)],use = "p")
#And then plot it with nice options 
ggcorrplot(corr_coef, 
           type = "lower",         # lower triangle of the matrix only
           hc.order = TRUE,        # variable sorted from highest to lowest
           outline.col = "white",  #Color options
           lab = TRUE)

```

## 4.2 Logistic regression

We will use the DHS survey at the individual level and construct a binary response variable indicating whether there was marriage before 15 years old (or "*girl's marriage*". We will try to understand which factor**S** may affect the probability to observe girl's marriage, and in particular whether, after controlling for the usual individual factors such as education, wealth and other socioeconomic factors, there is also an impact of environmental factors.   


```{r}
# We use the dhsdataMerge function to merge the survey data (individuals)
# with all the Geo-covariate extracted at the cluster level
DataDHS<-dhsdataMerge(DHS)

# We need to have a factor variable and not directly Before15 (that is numeric here)  
DataDHS$I_Before15 <- as.factor(DataDHS$Before15)

# Education is a factor variable
DataDHS$Education <- as.factor(DataDHS$Education)
# DataDHS <- DataDHS %>%                    # defining the reference category
#   mutate(Education = relevel(Education, "0-No"))
# 

# We change the unit of Aridity here 
DataDHS$Aridity2015 <- DataDHS$Aridity2015 * 10^8

# Defining the variables of the model
Y<-"I_Before15"               # Response variable
XCovars <- c(15, 17, 57:64)   # age+education+GIS

formula_string<- paste(Y, paste(colnames(DataDHS)[XCovars], collapse=" + "), sep="~")
print(paste(" Regression formula: ",formula_string))

```
### Results {-}
We use a logit model with several explanatory variables

```{r, results='asis'}
# Logistics Regression
glm.fit <- glm(formula_string, data = DataDHS, family = binomial)

# Nice printing of the results (using paper and knitr packages)
pretty_lm2 <- prettify(summary(glm.fit))
kable(pretty_lm2, digits = 3)

```


### Confusion Matrix {-}
```{r, results=TRUE }
library("regclass")
confusion_matrix(glm.fit)
```


### Visual representation of the logistic model{-} 


```{r visreg}
library(visreg)
library(ggpubr)

# Probabilities of married before 15 wrt 
p.age <- visreg(glm.fit, "Age", scale="response", rug=0,  # for rugs =2
       xlab="Age",
       ylab="P(Before15=1)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

p.education <- visreg(glm.fit, "Education", scale="response", rug=0,
       xlab="Education",
       ylab="P(Before15=1)", gg=TRUE) + 
  ylim(0,1) + theme_minimal() + 
 theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust=1,
                                   size=7))


p.aridity <- visreg(glm.fit, "Aridity2015", scale="response", rug=0,
       xlab="Aridity level (2015)",
       ylab="P(Before15=1)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

p.income <- visreg(glm.fit, "aIncome2013", scale="response", rug=0,
       xlab=" Estimated income (in $ 2013)",
       ylab="P(Before15=1)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()


figure <- ggarrange( p.age, p.education, p.aridity, p.income,
                    #labels = c("Edudation", "Age",  "Aridity (2015)", ""),
                    ncol = 2, nrow = 2)
figure
```


# 4.3 Random Forests  
 
 
```{r RF, cache = TRUE}
set.seed(888)               # set random seed so we can reproduce the result
myRandomForest<-randomForest(as.formula(formula_string),
                             data = DataDHS,
                             importance = TRUE,
                             maxnodes=25,
                             ntree=1000,
                             type="classification",
                             na.action = na.roughfix)
```

### Accuracy rate and confusion Matrix {-}

```{r, results = TRUE}
myRandomForest

```


### Variable importance plot {-}

```{r}
varImpPlot(myRandomForest, 
           type = 1,
           main =" Importance Plot for Random Forest Model")
```



